---
title: "MultivariateEDA_G4"
author: "MLDS Group 4"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.asp = 1)

# Check if libraries are installed, if not install them
if (!require(tidyverse)) install.packages("tidyve")
library(tidyverse)

if (!require(ggplot2)) install.packages("ggplot2")
library(ggplot2)

if (!require(corrplot)) install.packages("corrplot")
library(corrplot)

if (!require(dplyr)) install.packages("dplyr")
library(dplyr)

if (!require(GGally)) install.packages("GGally")
library(GGally)

if (!require(zoo)) install.packages("zoo")
library(zoo)

if (!require(patchwork)) install.packages("patchwork")
library(patchwork)

```

# 1. Load .csv data

```{r}
# Read in energy demand data
ed <- read_csv("energy_demand_uk.csv", show_col_types = FALSE) %>% select(-1)  

# Ensure date column is Date type
ed <- ed %>%
  mutate(date = as.Date(date))

```

# 2. Data Exploration
## Shape of data
```{r}
# dimensions of data
cat("Number of rows:", nrow(ed), "\n")
cat("Number of columns:", ncol(ed), "\n")
```

## Data summary
```{r}
cat("Summary of data")
summary(ed)
```

## Preview data
```{r}
print("Preview of data:")
head(ed)
```
# 3. Handle missing values
## Find which rows and columns have values missing
```{r}
# Check and print rows with missing values
missing_rows <- ed %>% filter(if_any(everything(), is.na))
cat("number of missing: rows: ", nrow(missing_rows), "\n")

# Check which columns have missing values
missing_cols <- names(row)[which(is.na(row))]
cat("Missing values in columns:", paste(missing_cols, collapse = ", "), "\n")

# Visualizing missing rows
print(missing_rows)

```

## Only 29/02 has missing values: min_temp, max_temp, rain_mm, wind_speed 
## forward fill missing values
```{r}
# Forward fill missing values
ed <- na.locf(ed)
```

# 4. Univariate EDA @Jas
# 5.1. Multivariate EDA: Original Features explorations
## Time Series plot of all variables and 30 day MA
```{r, fig.width=12, fig.height=6, message=FALSE, warning=FALSE}

# Remove cols not wanted
exclude_cols <- c("national_demand", "date")
columns_to_consider <- setdiff(names(ed), exclude_cols)

# Select what MA to plot
MA_TIME <- 30

# Create list of ggplots
plot_list <- lapply(columns_to_consider, function(col) {
  ggplot(ed, aes(x = date, y = .data[[col]])) +
    geom_line(color = "blue", linewidth = 1.2) +
    geom_line(aes(y = rollmean(.data[[col]], k=MA_TIME, fill = NA)), color = "red", linewidth = 1.2) +
    labs(title = col, x = "Date", y = col) +
    theme(axis.text.x = element_text(angle = 30, hjust = 1, size = 8))
})

# Combine all plots into a grid layout
combined_plot <- wrap_plots(plot_list) +
  plot_annotation(title = "Time Series with 30-Day Moving Average")

# Print the combined plot
print(combined_plot)

```

## Correlations to original Features only
```{r, message=FALSE, warning=FALSE}
# Calculate Spearman correlation for selected columns
correlations <- sapply(ed[columns_to_consider], 
                       function(column) cor(column, ed$national_demand, method = "spearman"))

print(correlations)
```

## Plot national_demand correlation to original Features only
```{r}
# Convert correlations to data frame for plotting
correlations_df <- data.frame(
  feature = names(correlations),
  correlation = as.numeric(correlations)
)

# Plot it
ggplot(correlations_df, aes(x = correlation, y = reorder(feature, correlation), fill = correlation > 0)) +
  geom_bar(stat = "identity", fill="salmon") +
  theme_minimal() +
  labs(
    title = "Correlation with National Demand (Original Features)",
    x = "Correlation",
    y = ""
  ) +
  theme(
    axis.text.y = element_text(size = 10),
    axis.title.y = element_text(size = 12),
    axis.title.x = element_text(size = 12),
    plot.title = element_text(size = 14, face = "bold"),
    legend.position = "none"
  )

```

## Plot national_demand against Original Features
```{r, fig.width=20, fig.height=12, message=FALSE, warning=FALSE}
# Create list of plots
plot_list <- lapply(columns_to_consider, function(col) {
  ggplot(ed, aes(x = .data[[col]], y = national_demand)) +
    geom_point(color = "black", alpha = 0.5) +  # Scatter plot
    geom_smooth(color = "red", method = "loess", se = FALSE) +  # Smooth trend line
    labs(title = paste(col),
         x = col,
         y = "National Demand") +
    theme_minimal()
})

# Combine all plots into a grid using patchwork
combined_plot <- wrap_plots(plot_list, ncol = 2) +
  plot_annotation(title = "National Demand vs Features")

# Show the combined plot
print(combined_plot)

```

# 5.2. Multivariate EDA: Transformations and Feature Engineering @Katja
# 5.3. Multivariate EDA: Confounding Effect - Solar Generation v. Temperature
## It can be seen that Solar generation correlates to demand, but is it causality?
Because temperature correlates strongly not only with demand, but also with 
solar_generation, it is a confounder. In fact, demand is seasonal and is not
affected by how much solar power is generated.
 
```{r, message=FALSE, warning=FALSE}
correlation <- cor(ed$solar_generation, ed$max_temp, method = "spearman")
print(correlation)
```

```{r, message=FALSE, warning=FALSE}

# Plot Solar Generation vs. Temperature
ggplot(ed, aes(x = solar_generation, y = max_temp)) +
geom_point(color = "black", alpha = 0.6) +
geom_smooth(method = "lm", color = "red", se = FALSE) +
labs(
  title = "Solar Generation vs. Max Temperature",
  x = "Solar Generation",
  y = "Max Temperature"
) +
theme_minimal()

```





# 5.4 daily_temp_variation
## calculate avg_temp, avg_temp_ma, daily_variation_avg_temp
```{r, message=FALSE, warning=FALSE}
# Calculate avg_temp
ed <- ed %>%
  mutate(avg_temp = (max_temp+min_temp)/2)

# Calculate avg_temp_ma_30
ed <- ed %>%
  mutate(avg_temp_ma = rollmean(avg_temp, k = 30, fill = NA, align = "right"))

# calculate daily_variation_avg_temp
ed <- ed %>%
  mutate(dail_variation_avg_temp = avg_temp - avg_temp_ma)
```

## Calculate daily variations correlation to demand - segmented by min_temperature
```{r, message=FALSE, warning=FALSE}

cold_threshold <- 5
hot_threshold <- 13

cold_days <- ed %>%
  filter(min_temp < cold_threshold)

mild_days <- ed %>%
  filter(min_temp > cold_threshold & min_temp < hot_threshold)

hot_days <- ed %>%
  filter(min_temp > hot_threshold)

# calculate correlation of demand v temperature by segment
correlation <- cor(cold_days$dail_variation_avg_temp, cold_days$national_demand, 
                   method="spearman", use = "complete.obs")
print(correlation)

correlation <- cor(mild_days$dail_variation_avg_temp, mild_days$national_demand, 
                   method="pearson", use = "complete.obs")
print(correlation)

correlation <- cor(hot_days$dail_variation_avg_temp, hot_days$national_demand, 
                   method="spearman", use = "complete.obs")
print(correlation)

```

## Plot daily variations correlation to demand - segmented by min_temperature
```{r, message=FALSE, warning=FALSE}
# Add segmentation flags to main data frame for plotting
ed <- ed %>%
  mutate(day_type = case_when(
    min_temp < cold_threshold ~ "cold_day",
    min_temp >= cold_threshold & min_temp < hot_threshold ~ "mild_day",
    TRUE ~ "hot_day"
  ))

# Plot
ggplot(ed, aes(x = dail_variation_avg_temp, y = national_demand, color = day_type)) +
  geom_point(alpha = 0.2) +  # Original points
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "National Demand vs Daily Variations in Average Temperature",
    x = "Daily Variations in Average Temperature",
    y = "National Demand"
  ) +
  scale_color_manual(values = c(
    "cold_day" = "blue", "mild_day" = "green", "hot_day" = "red")) +
  theme_minimal()

```

## During cold periods, daily temperature drops cause demand to increase
during mild and hot days, this behaviour is not seen