---
title: "Univariate Exploratory Data Analysis (EDA)"
output: html_document
---

```{r}
############## Loading the required Libary ############## 
# Load necessary libraries
library(ggplot2)
library(dplyr)
library(readr)
library(zoo)
library(forecast)
library(lubridate)
library(MASS)
#install.packages("tseries")  
library(tseries)

```

##Data Summary
---

```{r}
setwd("~/Desktop/Imperial/EDAV/EDAV Assessment 2")  #Change the path to your own path plz. 
energy_data <- read_csv("energy_demand_uk.csv")
head(energy_data)
summary(energy_data$national_demand)
```
##Checking Normality & All other types of distribution
---
```{r}
########################################################
##############  Checking Normaltiy        ##############
########################################################
#Using historgram
ggplot(energy_data, aes(x = national_demand)) +
  geom_histogram(binwidth = 1000, fill = "blue", alpha = 0.6, color = "black") +
  labs(title = "Distribution of National Energy Demand", x = "National Demand (MW)", y = "Count") +
  theme_minimal()
#Using boxplot
ggplot(energy_data, aes(y = national_demand)) +
  geom_boxplot(fill = "red", alpha = 0.5) +
  labs(title = "Boxplot of National Energy Demand", y = "National Demand (MW)") +
  theme_minimal()
#Using qq plot
ggplot(energy_data, aes(sample = national_demand)) +
  geom_qq() +
  geom_qq_line() +
  labs(title = "Q-Q Plot of National Energy Demand") +
  theme_minimal()

ggplot(energy_data, aes(x = date, y = national_demand)) +
  geom_line(color = "blue") +
  labs(title = "Time Series of National Energy Demand", x = "Date", y = "National Demand (MW)") +
  theme_minimal()


#Conclusion:
#The national demand data does not follow normal distribution
#So in the following steps, let's try some transformation, to see if the data can follow normal distribution after transformation


#Used Log transformation for national_demand:
#Using qq plot:
qqnorm(log(energy_data$national_demand))
qqline(log(energy_data$national_demand), col = "red")

#Fitting Gamma Distribution:
fit_gamma <- fitdistr(energy_data$national_demand, "gamma")
print(fit_gamma)  
shape_param <- fit_gamma$estimate["shape"]
rate_param <- fit_gamma$estimate["rate"]
hist(energy_data$national_demand, probability = TRUE, col = "lightblue", breaks = 30,
     main = "Gamma Fit on National Demand", xlab = "National Demand (MW)")
curve(dgamma(x, shape = shape_param, rate = rate_param), 
      add = TRUE, col = "red", lwd = 2)
ks.test(energy_data$national_demand, "pgamma", shape_param, rate_param)  #KS test 
# p-value (0.002129) from the KS test is very small (< 0.05), we reject H0. This concludes Gamma is not a good fit for the energy data

#Fitting Log Normal Distribution:
fit_lognormal <- fitdistr(energy_data$national_demand, "lognormal")
print(fit_lognormal)
meanlog <- fit_lognormal$estimate["meanlog"]
sdlog <- fit_lognormal$estimate["sdlog"]
ks.test(energy_data$national_demand, "plnorm", meanlog, sdlog) #KS test 
hist(energy_data$national_demand, probability = TRUE, col = "lightblue", breaks = 30,
     main = "Log-Normal Fit on National Demand", xlab = "National Demand (MW)")
curve(dlnorm(x, meanlog = meanlog, sdlog = sdlog), add = TRUE, col = "red", lwd = 2)
# p-value from the KS test is very small (< 0.05), we reject H0. This concludes Log-Normal is not a good fit for the energy data



```
##Conclusion:
##From the histrogram, the data appears to be right skewed, ie longer tail on the right. 
##From the qq plot, the data doesn't follow a normal distribution, based on the deviation from the extreme right and left side of the plot. 
##We did further testing on the log transformed of national demand, and from the qq plot we can see that it still doesn't fit normal distribution.
##Based on the right skewed feature of the data, we tried the following:
##1. Fitting gamma distribution, as the gamma distribution is only for positive values (X>0), and tend to be used for data that are right skewed. However, from the result for ##the KS test, we see that p-value (0.002129) from the KS test is very small (< 0.05), we reject H0. This concludes Gamma is not a good fit for the energy data

##2. Fitting Log Normal Distribution:
##Log-Normal distribution is also often used for postively skewed data.
##However from the KS test result, we reject H0. This concludes log normal is not a good fit for the energy data


##Box Plot By Month & Year
---
```{r}
#######################################################
##########         Box Plot by Month        ##########
#######################################################
# Ensure date column is in Date format
energy_data$date <- as.Date(energy_data$date)
energy_data$month <- factor(month(energy_data$date, label = TRUE), levels = month.abb)
ggplot(energy_data, aes(x = month, y = national_demand)) +
  geom_boxplot(fill = "lightblue", color = "black") +
  labs(title = "Distribution of National Energy Demand by Month",
       x = "Month", y = "National Demand (MW)") +
  theme_minimal()

#######################################################
##########        Box Plot by year          ##########
#######################################################

energy_data$date <- as.Date(energy_data$date)
energy_data$year <- factor(year(energy_data$date))
ggplot(energy_data, aes(x = year, y = national_demand)) +
  geom_boxplot(fill = "lightblue", color = "black") +
  labs(title = "Distribution of National Energy Demand by Year",
       x = "Year", y = "National Demand (MW)") +
```


```{r}

energy_data$date <- as.Date(energy_data$date)

# Define seasons
energy_data <- energy_data %>%
  mutate(season = case_when(
    month(date) %in% c(12, 1, 2) ~ "Winter",
    month(date) %in% c(3, 4, 5) ~ "Spring",
    month(date) %in% c(6, 7, 8) ~ "Summer",
    month(date) %in% c(9, 10, 11) ~ "Autumn"
  ))

# Compute seasonal mean and standard deviation
seasonal_stats <- energy_data %>%
  group_by(season) %>%
  summarise(mean_demand = mean(national_demand, na.rm = TRUE),
            sd_demand = sd(national_demand, na.rm = TRUE)) %>%
  mutate(season = factor(season, levels = c("Winter", "Spring", "Summer", "Autumn")))  # Ensure correct order

# Plot seasonal mean with error bars (standard deviation)
ggplot(seasonal_stats, aes(x = season, y = mean_demand)) +
  geom_bar(stat = "identity", fill = "skyblue", color = "black") +
  geom_errorbar(aes(ymin = mean_demand - sd_demand, ymax = mean_demand + sd_demand), width = 0.2, color = "red") +
  labs(title = "Mean and Standard Deviation of National Energy Demand by Season",
       x = "Season", y = "National Demand (MW)") +
  theme_minimal()



```
##The box is the mean of the energy demdn
##We can see in both winter and autumn, there is high demand for energy
## Winter, Spring and Autmn has the largest fluctuation of the demand (showsn by the red range)



##Time Series Analysis
---
```{r}
###################################      
######    Visualize the TS    #####
################################### 
ts_data <- ts(energy_data$national_demand, 
              start = c(2020, 1), 
              frequency = 365)  

ggplot(energy_data, aes(x = date, y = national_demand)) +
  geom_line(color = "blue") +
  labs(title = "Time Series of National Energy Demand",
       x = "Date", 
       y = "National Demand") +
  theme_minimal()


###################################      
###### Plotting Decomposed TS #####
###################################  
decomposed_ts <- decompose(ts_data, type = "multiplicative")
plot(decomposed_ts)
mtext("Time Series Decomposition: Trend, Seasonality, and Residuals", 
      side = 3, line = 1, outer = TRUE, cex = 1.5, font = 2)

###################################      
######## Yearly mean trend ######## 
###################################      
yearly_mean <- energy_data %>%
  group_by(year) %>%
  summarise(mean_demand = mean(national_demand))

ggplot(yearly_mean, aes(x = year, y = mean_demand)) +
  geom_line() +
  geom_point() +
  labs(title = "Yearly Mean National Energy Demand", 
       x = "Year", 
       y = "Mean Demand (MW)") +
  theme_minimal()

###################################      
#####   Check for Stationarity ####
################################### 
#Augmented Dickey-Fuller Test:
adf_test_result <- adf.test(ts_data)
print(adf_test_result)

#Given the p-value < 0.05, we reject H₀, and we conclude that time series is stationary.

###################################      
#####      Plotting MA         ####
###################################
energy_data <- energy_data %>%
  mutate(ma7 = zoo::rollmean(national_demand, k = 7, fill = NA),  # 7-day moving average
         ma30 = zoo::rollmean(national_demand, k = 30, fill = NA)) # 30-day moving average

ggplot(energy_data, aes(x = date)) +
  geom_line(aes(y = national_demand), color = "blue", alpha = 0.5) +
  geom_line(aes(y = ma7), color = "red", size = 1) +
  geom_line(aes(y = ma30), color = "green", size = 1) +
  labs(title = "Moving Averages of National Energy Demand",
       x = "Date", 
       y = "National Demand (MW)") +
  theme_minimal()

###################################      
#####  Plotting ACF & PACF     ####
###################################
acf(ts_data, main = "Autocorrelation Function (ACF) of National Demand")
pacf(ts_data, main = "Partial Autocorrelation Function (PACF) of National Demand")
adf.test(ts_data)  # Augmented Dickey-Fuller Test


```
## If we compare the over trend from the decomposition graph vs the trend of the yearly mean. The decomposition graph suggest stricly decreasing demand, while the yearly mean doesn't sugget this conclusion
## The Graph did suggest seasonality
## The noise graph looks completely random, where there is no patter. 

##MA Graph:
##From the original blue grpah -> we can see that there are a lot of noise, and sudden fluctuations (could be due to sudden change / requirement of energy)
##From the red plot -> smooth out some fluctuations
##From the green plot -> smooth out more fluctuations, and shows clear seasonality


## Fitting SARIMA
```{r pressure, echo=FALSE}
##Fitting SARIMA:
sarima_model <- auto.arima(ts_data, seasonal = TRUE, stepwise = FALSE, approximation = FALSE)
summary(sarima_model)
forecast_sarima <- forecast(sarima_model, h = 365)  # Forecast for next year
autoplot(forecast_sarima)
```






