---
title: "UK Energy Demand Analysis"
author: "MLDS Group 4"
output: html_document
date: "2025-03-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.asp = 1)

required_packages <- c("tidyverse", 
                      "zoo",        
                      "forecast",   
                      "lubridate",  
                      "MASS",       
                      "tseries",    
                      "patchwork")

# Load packages
for (pkg in required_packages) {
  if (!require(pkg, character.only = TRUE)) {
    install.packages(pkg)
    library(pkg, character.only = TRUE)
  }
}
```

# 1. Data Introduction and Cleaning

## Loading the Data

```{r}
energy_data <- read.csv("energy_demand_uk_1.csv")

# Ensure date column is Date type
energy_data$date <- as.Date(energy_data$date)

# Display data summary
cat("Data dimensions:", nrow(energy_data), "rows,", ncol(energy_data), "columns\n")
head(energy_data)
summary(energy_data)
```

## Handling Missing Values

```{r}
# Check and print rows with missing values
missing_rows <- energy_data %>% filter(if_any(everything(), is.na))
cat("Number of missing rows:", nrow(missing_rows), "\n")

# Identify which columns have missing values
if(nrow(missing_rows) > 0) {
  missing_cols <- colnames(missing_rows)[colSums(is.na(missing_rows)) > 0]
  cat("Columns with missing values:", paste(missing_cols, collapse=", "), "\n")
  
  # View the missing rows
  print(missing_rows)
  
  # Forward fill missing values
  energy_data <- na.locf(energy_data)
  
  # Verify missing values have been filled
  cat("Remaining missing values:", sum(is.na(energy_data)), "\n")
}
```

# 2. Univariate Exploratory Data Analysis

## Distribution of National Energy Demand

```{r}
# Histogram
ggplot(energy_data, aes(x = national_demand)) +
  geom_histogram(binwidth = 1000, fill = "blue", alpha = 0.6, color = "black") +
  labs(title = "Distribution of National Energy Demand", x = "National Demand (MW)", y = "Count") +
  theme_minimal()
```

```{r}
# Boxplot
ggplot(energy_data, aes(y = national_demand)) +
  geom_boxplot(fill = "red", alpha = 0.5) +
  labs(title = "Boxplot of National Energy Demand", y = "National Demand (MW)") +
  theme_minimal()
```

```{r}
# QQ plot
ggplot(energy_data, aes(sample = national_demand)) +
  geom_qq() +
  geom_qq_line() +
  labs(title = "Q-Q Plot of National Energy Demand") +
  theme_minimal()
```

```{r}
# Time series plot
ggplot(energy_data, aes(x = date, y = national_demand)) +
  geom_line(color = "blue") +
  labs(title = "Time Series of National Energy Demand", 
       x = "Date", y = "National Demand (MW)") +
  theme_minimal()
```

## Conclusion:
## From the histogram, the data appears to be right skewed, i.e., longer tail on the right. 
## From the QQ plot, the data doesn't follow a normal distribution, based on the deviation from the extreme right and left side of the plot. 
## The national demand data does not follow normal distribution.
## So in the following steps, let's try some transformation, to see if the data can follow normal distribution after transformation.

## Testing Different Distributions

```{r}
# QQ plot for log-transformed data
ggplot(energy_data, aes(sample = log(national_demand))) +
  geom_qq() +
  geom_qq_line(color = "red") +
  labs(title = "Q-Q Plot of Log-Transformed National Demand")
```

```{r}
# Fitting Gamma Distribution
fit_gamma <- fitdistr(energy_data$national_demand, "gamma")
print(fit_gamma)  
```
```{r}
shape_param <- fit_gamma$estimate["shape"]
rate_param <- fit_gamma$estimate["rate"]

# Plot histogram with gamma density using ggplot2 approach
ggplot(energy_data, aes(x = national_demand)) +
  geom_histogram(aes(y = ..density..), bins = 30, fill = "lightblue", color = "black") +
  stat_function(fun = dgamma, args = list(shape = shape_param, rate = rate_param),
                color = "red", linewidth = 1) +
  labs(title = "Gamma Fit on National Demand", x = "National Demand (MW)", y = "Density") +
  theme_minimal()
```

```{r}
# KS test for Gamma fit
ks_gamma <- ks.test(energy_data$national_demand, "pgamma", shape_param, rate_param)
print(ks_gamma)
```
```{r}
# Fitting Log Normal Distribution
fit_lognormal <- fitdistr(energy_data$national_demand, "lognormal")
print(fit_lognormal)
```
```{r}
meanlog <- fit_lognormal$estimate["meanlog"]
sdlog <- fit_lognormal$estimate["sdlog"]

# KS test for Log-Normal fit
ks_lognormal <- ks.test(energy_data$national_demand, "plnorm", meanlog, sdlog)
print(ks_lognormal)
```
```{r}
# Plot histogram with log-normal density using ggplot2 approach
ggplot(energy_data, aes(x = national_demand)) +
  geom_histogram(aes(y = ..density..), bins = 30, fill = "lightblue", color = "black") +
  stat_function(fun = dlnorm, args = list(meanlog = meanlog, sdlog = sdlog),
                color = "red", linewidth = 1) +
  labs(title = "Log-Normal Fit on National Demand", x = "National Demand (MW)", y = "Density") +
  theme_minimal()
```

## Conclusion:
## From the histogram, the data appears to be right skewed, i.e., longer tail on the right. 
## From the QQ plot, the data doesn't follow a normal distribution, based on the deviation from the extreme right and left side of the plot. 
## We did further testing on the log transformed of national demand, and from the QQ plot we can see that it still doesn't fit normal distribution.
## Based on the right skewed feature of the data, we tried the following:
## 1. Fitting gamma distribution, as the gamma distribution is only for positive values (X>0), and tend to be used for data that are right skewed. However, from the result for the KS test, we see that p-value (0.002129) from the KS test is very small (< 0.05), we reject H0. This concludes Gamma is not a good fit for the energy data.
## 2. Fitting Log Normal Distribution:
## Log-Normal distribution is also often used for positively skewed data.
## However from the KS test result, we reject H0. This concludes log normal is not a good fit for the energy data.

## Distribution by Month and Year

```{r}
# Boxplot by Month
energy_data$month <- factor(month(energy_data$date, label = TRUE), levels = month.abb)
ggplot(energy_data, aes(x = month, y = national_demand)) +
  geom_boxplot(fill = "lightblue", color = "black") +
  labs(title = "Distribution of National Energy Demand by Month",
       x = "Month", y = "National Demand (MW)") +
  theme_minimal()
```
```{r}
# Boxplot by Year
energy_data$year <- factor(year(energy_data$date))
ggplot(energy_data, aes(x = year, y = national_demand)) +
  geom_boxplot(fill = "lightblue", color = "black") +
  labs(title = "Distribution of National Energy Demand by Year",
       x = "Year", y = "National Demand (MW)") +
  theme_minimal()
```

## Seasonal Analysis

```{r}
# Define seasons
energy_data <- energy_data %>%
  mutate(season = case_when(
    month(date) %in% c(12, 1, 2) ~ "Winter",
    month(date) %in% c(3, 4, 5) ~ "Spring",
    month(date) %in% c(6, 7, 8) ~ "Summer",
    month(date) %in% c(9, 10, 11) ~ "Autumn"
  ))

# Ensure correct order of seasons
energy_data$season <- factor(energy_data$season, 
                            levels = c("Winter", "Spring", "Summer", "Autumn"))

# Compute seasonal mean and standard deviation
seasonal_stats <- energy_data %>%
  group_by(season) %>%
  summarise(mean_demand = mean(national_demand, na.rm = TRUE),
            sd_demand = sd(national_demand, na.rm = TRUE))

# Plot seasonal mean with error bars (standard deviation)
ggplot(seasonal_stats, aes(x = season, y = mean_demand)) +
  geom_bar(stat = "identity", fill = "skyblue", color = "black") +
  geom_errorbar(aes(ymin = mean_demand - sd_demand, ymax = mean_demand + sd_demand), 
                width = 0.2, color = "red") +
  labs(title = "Mean and Standard Deviation of National Energy Demand by Season",
       x = "Season", y = "National Demand (MW)") +
  theme_minimal()
```

## Conclusion:
## The bar plot represents the mean of the energy demand.
## We can see in both winter and autumn, there is high demand for energy.
## Winter, Spring and Autumn have the largest fluctuation of the demand (shown by the red error bars).

## Time Series Analysis

```{r}
# Create time series object
ts_data <- ts(energy_data$national_demand, 
              start = c(2020, 1), 
              frequency = 365)  

# Time series plot with ggplot2
ggplot(energy_data, aes(x = date, y = national_demand)) +
  geom_line(color = "blue") +
  labs(title = "Time Series of National Energy Demand",
       x = "Date", 
       y = "National Demand") +
  theme_minimal()
```
```{r}
# Decompose time series
decomposed_ts <- decompose(ts_data, type = "multiplicative")

# Plot the decomposition
plot(decomposed_ts)
```
```{r}
# Yearly mean trend
yearly_mean <- energy_data %>%
  group_by(year) %>%
  summarise(mean_demand = mean(national_demand))

ggplot(yearly_mean, aes(x = year, y = mean_demand)) +
  geom_line(group = 1) +
  geom_point() +
  labs(title = "Yearly Mean National Energy Demand", 
       x = "Year", 
       y = "Mean Demand (MW)") +
  theme_minimal()
```
```{r}
# Check for Stationarity with Augmented Dickey-Fuller Test
adf_test_result <- adf.test(ts_data)
print(adf_test_result)
```
```{r}
# Moving averages
energy_data <- energy_data %>%
  mutate(ma7 = zoo::rollmean(national_demand, k = 7, fill = NA),  # 7-day moving average
         ma30 = zoo::rollmean(national_demand, k = 30, fill = NA)) # 30-day moving average

ggplot(energy_data, aes(x = date)) +
  geom_line(aes(y = national_demand), color = "blue", alpha = 0.5) +
  geom_line(aes(y = ma7), color = "red", size = 1) +
  geom_line(aes(y = ma30), color = "green", size = 1) +
  labs(title = "Moving Averages of National Energy Demand",
       x = "Date", 
       y = "National Demand (MW)") +
  theme_minimal()
```
```{r}
# ACF and PACF
acf(ts_data, main = "Autocorrelation Function (ACF) of National Demand")
```
```{r}
pacf(ts_data, main = "Partial Autocorrelation Function (PACF) of National Demand")
```

## Conclusion:
## If we compare the overall trend from the decomposition graph vs the trend of the yearly mean, the decomposition graph suggests strictly decreasing demand, while the yearly mean doesn't suggest this conclusion.
## The decomposition graph did suggest seasonality.
## The noise graph looks completely random, where there is no pattern.
## Given the p-value < 0.05 from the ADF test, we reject H₀, and we conclude that the time series is stationary.

## Moving Average Analysis:
## From the original blue graph -> we can see that there are a lot of noise, and sudden fluctuations (could be due to sudden change / requirement of energy).
## From the red plot (7-day MA) -> smooths out some fluctuations.
## From the green plot (30-day MA) -> smooths out more fluctuations, and shows clear seasonality.

## SARIMA Model Fitting

```{r}
# Fit SARIMA model
sarima_model <- auto.arima(ts_data, seasonal = TRUE, 
                          stepwise = FALSE, approximation = FALSE)
summary(sarima_model)

# Forecast
forecast_sarima <- forecast(sarima_model, h = 365)  # Forecast for next year
autoplot(forecast_sarima)
```

# 3. Multivariate Exploratory Data Analysis

## Time Series of All Variables

```{r, fig.width=12, fig.height=15, message=FALSE, warning=FALSE}
# Variables to exclude
exclude_cols <- c("national_demand", "date", "month", "year", "season", "ma7", "ma30")

# Identify columns to analyze
columns_to_consider <- setdiff(names(energy_data), exclude_cols)

# MA time window
MA_TIME <- 30

# Create list of ggplots
plot_list <- lapply(columns_to_consider, function(col) {
  ggplot(energy_data, aes(x = date, y = .data[[col]])) +
    geom_line(color = "blue", linewidth = 0.8) +
    geom_line(aes(y = rollmean(.data[[col]], k=MA_TIME, fill = NA)), 
              color = "red", linewidth = 0.8) +
    labs(title = col, x = "Date", y = col) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 30, hjust = 1, size = 8))
})

# Combine all plots into a grid layout
combined_plot <- wrap_plots(plot_list, ncol = 2) +
  plot_annotation(title = "Time Series with 30-Day Moving Average")

# Print the combined plot
print(combined_plot)
```

## Correlation Analysis

```{r}
# Calculate Spearman correlation for original features
correlations <- sapply(energy_data[columns_to_consider], 
                       function(column) cor(column, energy_data$national_demand, 
                                           method = "spearman", 
                                           use = "complete.obs"))

# Display correlations
print(correlations)
```
```{r}
# Convert correlations to data frame for plotting
correlations_df <- data.frame(
  feature = names(correlations),
  correlation = as.numeric(correlations)
)

# Plot correlations
ggplot(correlations_df, aes(x = correlation, y = reorder(feature, correlation))) +
  geom_bar(stat = "identity", fill = "salmon") +
  theme_minimal() +
  labs(
    title = "Correlation with National Demand (Original Features)",
    x = "Correlation",
    y = ""
  ) +
  theme(
    axis.text.y = element_text(size = 10),
    axis.title.y = element_text(size = 12),
    axis.title.x = element_text(size = 12),
    plot.title = element_text(size = 14, face = "bold"),
    legend.position = "none"
  )
```
## Plot national_demand against Original Features
```{r, fig.width=20, fig.height=12, message=FALSE, warning=FALSE}
# Create list of plots
plot_list <- lapply(columns_to_consider, function(col) {
  ggplot(ed, aes(x = .data[[col]], y = national_demand)) +
    geom_point(color = "black", alpha = 0.5) +  # Scatter plot
    geom_smooth(color = "red", method = "loess", se = FALSE) +  # Smooth trend line
    labs(title = paste(col),
         x = col,
         y = "National Demand") +
    theme_minimal()
})

# Combine all plots into a grid using patchwork
combined_plot <- wrap_plots(plot_list, ncol = 2) +
  plot_annotation(title = "National Demand vs Features")

# Show the combined plot
print(combined_plot)

```
## Feature Engineering

```{r}
# Add engineered features
energy_data <- energy_data %>%
  mutate(
    avg_temp = (min_temp + max_temp) / 2,
    day_of_year = yday(date),
    annual_cos = cos(2 * pi * day_of_year / 365),
    HDD = pmax(15.5 - avg_temp, 0),  # Heating Degree Days (base 15.5°C)
    
    # Lag features
    demand_lag1 = lag(national_demand, 1),
    demand_lag2 = lag(national_demand, 2),
    demand_lag7 = lag(national_demand, 7),
    demand_lag365 = lag(national_demand, 365),
    
    # Moving averages
    demand_ma7 = if("ma7" %in% names(energy_data)) ma7 else 
                 rollmean(national_demand, k = 7, fill = NA, align = "right"),
    demand_ma30 = if("ma30" %in% names(energy_data)) ma30 else 
                  rollmean(national_demand, k = 30, fill = NA, align = "right"),
    demand_ma365 = rollmean(national_demand, k = 365, fill = NA, align = "right"),
    
    # Renewable energy ratio
    renewable_log_diff = log(wind_generation + 1) - log(solar_generation + 1)
  )

# Identify original and engineered features
original_features <- c("wind_generation", "solar_generation", "min_temp", 
                     "max_temp", "rain_mm", "wind_speed", "average_price_daily")

# Get all numeric column names
all_numeric_names <- names(select_if(energy_data, is.numeric))

# Create engineered features list (excluding national_demand and original features)
engineered_features <- setdiff(all_numeric_names, 
                             c("national_demand", original_features))

# Calculate correlations with national demand
numeric_cols <- energy_data %>% 
  select_if(is.numeric) %>%
  select_at(vars(-national_demand))

all_correlations <- data.frame(
  term = names(numeric_cols),
  correlation = sapply(numeric_cols, function(x) 
                      cor(x, energy_data$national_demand, use = "pairwise.complete.obs"))
) %>%
  # Add feature type
  mutate(feature_type = case_when(
    term %in% original_features ~ "Original",
    term %in% engineered_features ~ "Engineered",
    TRUE ~ "Other"
  ))

# Filter for strong correlations (|r| > 0.5)
strong_correlations <- all_correlations %>%
  filter(!is.na(correlation) & abs(correlation) > 0.5) %>%
  arrange(desc(abs(correlation)))

# Plot strong correlations
ggplot(strong_correlations, 
       aes(x = reorder(term, abs(correlation)), 
           y = correlation, 
           fill = feature_type)) +
  geom_col() +
  coord_flip() +
  labs(title = "Strong Correlations with National Demand", 
       x = "Features", 
       y = "Correlation",
       fill = "Feature Type") +
  theme_minimal() +
  scale_fill_brewer(palette = "Set2")
```

## Partial Dependence Plots

```{r, fig.width=12, fig.height=10, message=FALSE, warning=FALSE}
# Select key variables for partial dependence
demand_dependence_vars <- c("national_demand", 
                           "annual_cos", 
                           "HDD", 
                           "demand_lag1", 
                           "demand_lag7", 
                           "demand_lag365", 
                           "renewable_log_diff")

# Prepare data
plot_data <- energy_data %>%
  dplyr::select(all_of(demand_dependence_vars)) %>%
  na.omit()

create_partial_dependence_plot <- function(feature) {
  ggplot(plot_data, aes_string(x = feature, y = "national_demand")) +
    geom_point(alpha = 0.3) +
    geom_smooth(method = "loess", color = "red", se = TRUE) +
    labs(title = paste("Partial Dependence Plot:", feature),
         x = feature,
         y = "National Demand") +
    theme_minimal()
}

partial_plots <- lapply(setdiff(demand_dependence_vars, "national_demand"), 
                      create_partial_dependence_plot)
combined_partial_plots <- wrap_plots(partial_plots, ncol = 2)
print(combined_partial_plots)
```

## Confounding Effect: Solar Generation vs. Temperature

## It can be seen that Solar generation correlates to demand, but is it causality?
## Because temperature correlates strongly not only with demand, but also with 
## solar_generation, it is a confounder. In fact, demand is seasonal and is not
## affected by how much solar power is generated.

```{r, message=FALSE, warning=FALSE}
# Calculate correlation between solar generation and max temperature
solar_temp_corr <- cor(energy_data$solar_generation, energy_data$max_temp, 
                     method = "spearman", use = "complete.obs")
cat("Correlation between solar generation and max temperature:", solar_temp_corr, "\n")
```
```{r, message=FALSE, warning=FALSE}
# Plot Solar Generation vs. Temperature
ggplot(energy_data, aes(x = solar_generation, y = max_temp)) +
  geom_point(color = "black", alpha = 0.6) +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(
    title = "Solar Generation vs. Max Temperature",
    x = "Solar Generation (MW)",
    y = "Max Temperature (°C)"
  ) +
  theme_minimal()
```

## Daily Temperature Variation Effect

```{r, message=FALSE, warning=FALSE}
# Calculate avg_temp_ma_30 (if not already created)
if(!"avg_temp_ma" %in% names(energy_data)) {
  energy_data <- energy_data %>%
    mutate(avg_temp_ma = rollmean(avg_temp, k = 30, fill = NA, align = "right"))
}

# Calculate daily temperature variation
if(!"dail_variation_avg_temp" %in% names(energy_data)) {
  energy_data <- energy_data %>%
    mutate(dail_variation_avg_temp = avg_temp - avg_temp_ma)
}

# Define temperature thresholds
cold_threshold <- 5
hot_threshold <- 13

# Create temperature segments
energy_data <- energy_data %>%
  mutate(day_type = case_when(
    min_temp < cold_threshold ~ "cold_day",
    min_temp >= cold_threshold & min_temp < hot_threshold ~ "mild_day",
    TRUE ~ "hot_day"
  ))

# Calculate correlations by segment
cold_days <- energy_data %>% filter(min_temp < cold_threshold)
mild_days <- energy_data %>% filter(min_temp >= cold_threshold & min_temp < hot_threshold)
hot_days <- energy_data %>% filter(min_temp >= hot_threshold)

# Calculate correlations of demand vs temperature variation by segment
cat("Correlation in cold days:", 
    cor(cold_days$dail_variation_avg_temp, cold_days$national_demand, 
        method="spearman", use="complete.obs"), "\n")

cat("Correlation in mild days:", 
    cor(mild_days$dail_variation_avg_temp, mild_days$national_demand, 
        method="pearson", use="complete.obs"), "\n")

cat("Correlation in hot days:", 
    cor(hot_days$dail_variation_avg_temp, hot_days$national_demand, 
        method="spearman", use="complete.obs"), "\n")

# Plot daily variations correlation to demand by temperature segment
ggplot(energy_data, aes(x = dail_variation_avg_temp, y = national_demand, color = day_type)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "National Demand vs Daily Variations in Average Temperature",
    x = "Daily Variations in Average Temperature (°C)",
    y = "National Demand (MW)"
  ) +
  scale_color_manual(values = c(
    "cold_day" = "blue", "mild_day" = "green", "hot_day" = "red"),
    name = "Temperature\nSegment") +
  theme_minimal()
```

## During cold periods, daily temperature drops cause demand to increase. During mild and hot days, this behavior is not seen.